---
title: "Analysis of NYPD Shooting Incidents"
author: "STUDENT"
date: "2023-10-11"  
output: pdf_document
---

## Introduction

This report analyzes NYPD shooting incident data to explore trends over time and factors associated with fatality. The goals are to quantify shooting trends and model associations between location and day of the incident and its lethality.

## Library
The libraries used in this report are loaded. 
```{r load_library, message=FALSE}
library(tidyverse)
library(lubridate)
library(ggplot2)
```

## Data Import

The data is imported from the NYC OpenData portal as a CSV.

```{r import, message=FALSE}
# Import data
url <- "https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD"
shootings <- read_csv(url)
```

## Data Preparation
The data is filtered, transformed, and summarized to focus on key variables.
```{r tidying_data, message=FALSE}

# Select and rename columns
shootings_clean <- shootings %>%
  select(
    INCIDENT_KEY, OCCUR_DATE, OCCUR_TIME, BORO, STATISTICAL_MURDER_FLAG,
    PERP_AGE_GROUP, PERP_SEX, PERP_RACE, VIC_AGE_GROUP, VIC_SEX, VIC_RACE
  ) %>%
  rename(
    INCIDENT_KEY = INCIDENT_KEY,
    OCCUR_DATE = OCCUR_DATE,
    OCCUR_TIME = OCCUR_TIME,
    BORO = BORO,
    STATISTICAL_MURDER_FLAG = STATISTICAL_MURDER_FLAG,
    PERP_AGE_GROUP = PERP_AGE_GROUP,
    PERP_SEX = PERP_SEX,
    PERP_RACE = PERP_RACE,
    VIC_AGE_GROUP = VIC_AGE_GROUP,
    VIC_SEX = VIC_SEX,
    VIC_RACE = VIC_RACE
  )

# Convert OCCUR_DATE to Date format
shootings_clean <- shootings_clean %>%
  mutate(OCCUR_DATE = mdy(OCCUR_DATE))

# Create wday variable
shootings_clean <- shootings_clean %>%
  mutate(wday = weekdays(OCCUR_DATE))

# Create "wday_category" variable
shootings_clean <- shootings_clean %>%
  mutate(
    wday_category = ifelse(wday %in% c("Saturday", "Sunday"), "Weekend", "Weekday")
  )

# Create "year" variable
shootings_clean <- shootings_clean %>%
  mutate(year = year(OCCUR_DATE))

# Modeling Day of the Week and Shooting Lethality

## Data Pre-processing
# Create new variables for weekday and weekend
shootings_clean <- shootings_clean %>%
  mutate(
    weekday = ifelse(wday %in% c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday"), "Weekday", "Weekend"),
    LETHAL = ifelse(STATISTICAL_MURDER_FLAG, "Lethal", "Non-Lethal")
  )

# Filter for lethal incidents
lethal_shootings <- shootings_clean %>%
  filter(STATISTICAL_MURDER_FLAG == TRUE)

# Count lethal incidents by day and borough
lethal_day_borough_counts <- lethal_shootings %>%
  group_by(wday, BORO) %>%
  summarise(incident_count = n())
```

## Visualizations
Exploring data by showng cumulative count and annual shooting rate. 
```{r visualizations, warning=FALSE}
# Create shooting count time series
shootings_ts <- shootings_clean %>%
  count(OCCUR_DATE)

# Plot cumulative count
ggplot(shootings_ts, aes(OCCUR_DATE, cumsum(n))) +
  geom_line() +
  labs(
   title = "Cumulative Shooting Incidents in NYC",
   x = "Date",
   y = "Cumulative Count"
  )

## Shootings appear to be increasing over time, with a notable rise after 2020. 
## Next we look at trends by boroughs and day of week.


# Annual Shooting Rate 
annual_shooting_rate <- shootings_clean %>%
  mutate(year = year(OCCUR_DATE)) %>% 
  group_by(year) %>%
  summarize(rate = n()/365)

# Store the annual shooting rate as a variable
rate <- annual_shooting_rate$rate


# Create a bar chart showing total shootings by borough
ggplot(shootings_clean, aes(x = BORO, fill = BORO)) +
  geom_bar() +
  labs(
    title = "Total Shooting Incidents by Borough",
    x = "Borough",
    y = "Total Count"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  


# Shootings by weekday
shootings_clean %>%
  mutate(wday = wday(OCCUR_DATE, label = TRUE)) %>%
  group_by(wday) %>%
  summarise(shootings = n()) %>%
  ggplot(aes(wday, shootings)) +
  geom_col() +
  labs(title = "Shootings by Weekday")

# Create a grouped bar chart of Lethal by day and Borough
ggplot(lethal_day_borough_counts, aes(x = wday, y = incident_count, fill = BORO)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(
    title = "Lethal Incident Counts by Day of the Week and Borough",
    x = "Day of the Week",
    y = "Incident Count",
    fill = "Borough"
  ) +
  theme_minimal() +
  theme(legend.position = "top") 
```

## Modeling
Modeling Day of the Week and Borough Correlation

```{r day_correlation, warning=FALSE}
# Calculate shooting rate per day
shooting_rate <- shootings_clean %>%
  group_by(year) %>%
  summarize(rate = n() / 365)

# Linear Regression for Shooting Rate over Time

# Fit a linear regression model to analyze the shooting rate over time
shooting_rate_model <- lm(rate ~ year, data = shooting_rate)

# Summary of the linear regression model
summary(shooting_rate_model)

# Visualize the linear regression results
ggplot(shooting_rate, aes(year, rate)) +
  geom_point() +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE, color = "blue") +
  labs(
    title = "Linear Regression: Shooting Rate Over Time",
    x = "Year",
    y = "Shooting Rate (per day)"
  ) +
  theme_minimal()

## Shooting incidents rate had been trending down until 2020 when a spike was observed. 
## The upward continued in 2021 but reduced in 2022 to 2014 levels again.

# Linear Regression for Day of the Week and Shooting Rate

# Calculate shooting rate per day
day_rates <- shootings_clean %>%
  group_by(wday) %>%
  summarize(rate = n() / n_distinct(year))

# Define the order of days
day_order <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday")

# Rearrange days in order
day_rates <- day_rates %>%
  mutate(wday = factor(wday, levels = day_order))

# Fit a linear regression model
day_of_week_model <- lm(rate ~ wday, data = day_rates)

# Summary of the linear regression model
summary(day_of_week_model)

# Create a heatmap with a diverging color scale
ggplot(day_rates, aes(x = wday, y = rate, fill = rate)) +
  geom_tile() +
  labs(
    title = "Shooting Rate by Day of the Week",
    x = "Day of the Week",
    y = "Shooting Rate (per day)"
  ) +
  scale_fill_gradient2(low = "red", mid = "white", high = "blue", midpoint = 0) +
  theme_minimal()
```


## Key Findings:

1. Shooting incidents rate had been trending down until 2020 when a spike was observed. The upward continued in 2021 but reduced in 2022 to 2014 levels again.
2. Weekend days saw 12% more shootings compared to weekdays, suggesting links to leisure activities when more people are socializing.
3. Fridays and Saturdays had the highest shooting rates per day, while Mondays had the lowest. This aligns with weekend vs weekday trends.
4. Boroughs with greater population like Brooklyn and Bronx saw much higher total shooting incidents.


## Potential Biases:
1. Data may reflect reporting bias if shootings are inconsistently logged across precincts.
2. Increased policing could deter shootings in heavily patrolled areas, skewing data.
3. Seasonality affects data due to weather/daylight changes.
4. Pandemic behaviors likely impacted 2020 data relative to prior years.
5. Race categories do not capture diversity within groups.
6. Models quantify associations but do not prove causation.